---
title: "Lab 2: Exploratory Data Analysis – Student Exercises"
author: "Student Name:"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 2 – Student Exercises
**Topic:** Exploratory Data Analysis and Descriptive Statistics  
**Dataset:** `mtcars` (built-in R dataset)

---

## Scenario

You are working as a **junior data analyst** in an automotive research company.
Your task is to explore the `mtcars` dataset to identify patterns related to
**fuel efficiency**, **engine characteristics**, and **transmission types**.

You must answer each question using **R code** and provide **written interpretations**
of your results.

---

## Core Exercise Questions (To Be Done in the Lab)

### Question 1: Understanding the Dataset

1. Load the `mtcars` dataset.
mtcars
2. Display the first few rows of the dataset.
head(mtcars)
3. Identify which variables are **numerical** and which should be treated as **categorical**.
numerical
MPG,DISP,HP,DRAT,wt,qsec
categorical
cyl,vs,am,gear

4. Convert `cyl` (number of cylinders) and `am` (transmission type) into factors.
```{r}
# Write your code here

mtcars$cyl <- factor(mtcars$cyl)

mtcars$am <- factor(mtcars$am)
levels(mtcars$am) <- c("Automatic", "Manual")

```

Write a short explanation describing the variables in the dataset.

The variables were passed from numericals to factors
factors have levels like category labels (1 for automatic, 0 for manual etc)


---

### Question 2: Descriptive Statistics (Numerical Variable)

Fuel efficiency (`mpg`) is an important numerical variable.

1. Calculate the **mean**, **median**, and **standard deviation** of `mpg`.

2. Explain what each statistic tells you about fuel efficiency.

```{r}
# Write your code here

mean(mtcars$mpg)
median(mtcars$mpg)
sd(mtcars$mpg)
```


mean is the average across all cars (affected by extreme values)
Median is middle value when all cars are orderd
standard variation is the variability, high sd high variation of levels of mpg

---

### Question 3: Visualising Fuel Efficiency

1. Create a **histogram** of `mpg`.
2. Add appropriate axis labels and a title.



```{r}
# Write your code here


hist(mtcars$mpg,
     main = "Histogram of Fuel Efficiency (MPG)",
     xlab = "Miles per Gallon (MPG)",
     ylab = "Frequency",
     col = "lightblue",
     border = "black")

```

Write a short paragraph interpreting the shape and spread of the distribution.

---

### Question 4: Comparing Groups Using a Boxplot

1. Create a **boxplot** comparing `mpg` for cars with automatic and manual transmission.
2. Add appropriate labels.

```{r}
# Write your code here

boxplot(mpg ~ am,
        data = mtcars,
        main = "Fuel Efficiency (MPG) by Transmission Type",
        xlab = "Transmission Type",
        ylab = "Miles per Gallon (MPG)",
        col = c("lightgray", "lightblue"))

```

Explain what the boxplot tells you about fuel efficiency and transmission type.

---

### Question 5: Exploring Categorical Data

1. Create a **bar plot** showing the number of cars for each value of `cyl`.
2. Add a suitable title and axis labels.

```{r}
# Write your code here
barplot(table(mtcars$cyl),
        main = "Number of Cars by Cylinder Count",
        xlab = "Number of Cylinders",
        ylab = "Number of Cars",
        col = "lightblue",
        border = "black")

```

Explain what this graph shows about the composition of the dataset.

---

## Submission Notes

- Ensure all code runs without errors.
- Include **both R code and written explanations**.
- Save and upload this file as instructed on Blackboard.
